---
title: "Longitudinal Heatmap Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Heatmap Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This tutorial uses a real public RNA-seq time-course from the Bioconductor
`timecoursedata` package: sorghum leaf samples collected across weekly
drought/control conditions.

If the data package is not installed:

```{r install-timecoursedata, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("timecoursedata")
```

## 1. Load A Public Time-Course

```{r load, eval=FALSE}
library(HeatmapPlot)

example <- load_real_timecourse_example(
  dataset = "sorghum_leaf",
  genotype = "BT642",
  min_week = 3,
  top_n_features = 1000
)

dim(example$mat)
head(example$meta[, c(example$sample_col, example$time_col, example$group_col)])
```

The returned object is plain R data:

- `example$mat`: numeric feature-by-sample matrix.
- `example$meta`: one row per sample.
- `example$sample_col`, `example$time_col`, `example$group_col`: column names to
  pass into the plotting functions.

## 2. Choose The Number Of Modules

`choose_k()` calculates a simple within-cluster-sum-of-squares elbow diagnostic.
Use it as a first pass, then choose a biologically interpretable value.

```{r choose-k, fig.width=6, fig.height=4, eval=FALSE}
diag <- choose_k(
  zscore_rows(example$mat),
  k_range = 2:10,
  method = "kmeans"
)

diag
attr(diag, "suggested_k")
```

## 3. Cluster Temporal Patterns

```{r cluster, eval=FALSE}
obj <- longitudinal_cluster(
  example$mat,
  example$meta,
  sample_col = example$sample_col,
  time_col = example$time_col,
  group_col = example$group_col,
  k = attr(diag, "suggested_k"),
  method = "kmeans",
  aggregate_replicates = TRUE,
  stability = TRUE,
  n_resample = 20
)

obj
```

Clusters are reordered by their temporal peak, so early-high modules appear
before late-high modules.

## 4. Draw The Heatmap

```{r heatmap, fig.width=8, fig.height=10, eval=FALSE}
longitudinal_heatmap(
  obj,
  annotation_cols = c(example$group_col, example$time_col),
  top_n_label = 3,
  title = "Sorghum leaf drought time course"
)
```

To save a PDF:

```{r save-heatmap, eval=FALSE}
longitudinal_heatmap(
  obj,
  annotation_cols = c(example$group_col, example$time_col),
  top_n_label = 3,
  file = "sorghum_leaf_heatmap.pdf",
  width = 9,
  height = 12
)
```

## 5. Plot Module Trajectories

```{r lineplot, fig.width=7, fig.height=8, eval=FALSE}
pattern_lineplot(obj)
```

## 6. Export Tables

```{r write, eval=FALSE}
write_memberships(obj, prefix = "sorghum_leaf")
```

This writes:

- `sorghum_leaf_memberships.txt`
- `sorghum_leaf_centroids.txt`
- `sorghum_leaf_stability.txt`

## Other Public Examples

```{r available, eval=FALSE}
available_real_timecourse_datasets()
```

For the mouse influenza microarray dataset:

```{r influenza, eval=FALSE}
flu <- load_real_timecourse_example(
  dataset = "influenza_mouse",
  group = c("C", "K", "M"),
  top_n_features = 1000
)
```
