---
title: "Longitudinal Heatmap Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal Heatmap Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`ModuleViz` clusters feature trajectories from longitudinal omics experiments
and turns them into ordered temporal heatmaps, module trajectory plots, and
exportable membership tables. The workflow is useful for RNA-seq, ATAC-seq,
Cut&Run, or other feature-by-sample matrices where columns represent ordered
time points, treatment conditions, or both.

This vignette uses the small bundled `heatmap_example` dataset so the tutorial
runs without network access.

## 1. Load The Example Time-Course

```{r load}
library(ModuleViz)
data("heatmap_example")

mat <- heatmap_example$mat
meta <- heatmap_example$meta

dim(mat)
head(meta)
```

The returned object is plain R data:

- `mat`: numeric feature-by-sample matrix.
- `meta`: one row per sample, with sample ID, condition, time, and replicate.

## 2. Choose The Number Of Modules

`choose_k()` calculates a simple within-cluster-sum-of-squares elbow diagnostic.
Use it as a first pass, then choose a biologically interpretable value.

```{r choose-k}
diag <- choose_k(
  zscore_rows(mat),
  k_range = 2:5,
  method = "kmeans"
)

diag
attr(diag, "suggested_k")
```

## 3. Cluster Temporal Patterns

```{r cluster}
obj <- longitudinal_cluster(
  mat,
  meta,
  sample_col = "SampleID",
  time_col = "TimeHr",
  group_col = "Condition",
  k = 4,
  method = "kmeans",
  aggregate_replicates = TRUE,
  stability = TRUE,
  n_resample = 10
)

obj
```

Clusters are reordered by their temporal peak, so early-high modules appear
before late-high modules.

## 4. Draw The Heatmap

```{r heatmap, fig.width=8, fig.height=8}
longitudinal_heatmap(
  obj,
  annotation_cols = c("Condition", "TimeHr"),
  top_n_label = 3,
  title = "Example longitudinal heatmap"
)
```

To save a PDF:

```{r save-heatmap, eval=FALSE}
longitudinal_heatmap(
  obj,
  annotation_cols = c("Condition", "TimeHr"),
  top_n_label = 3,
  file = "example_heatmap.pdf",
  width = 9,
  height = 12
)
```

## 5. Plot Module Trajectories

```{r lineplot, fig.width=7, fig.height=7}
pattern_lineplot(obj)
```

## 6. Export Tables

```{r write}
out_dir <- tempdir()
write_memberships(obj, prefix = file.path(out_dir, "heatmap_example"))
```

This writes:

- `heatmap_example_memberships.txt`
- `heatmap_example_centroids.txt`
- `heatmap_example_stability.txt`

## Real Public Data

The package also includes a helper for public datasets in Bioconductor
`timecoursedata`. These examples are optional because they require installing
the experiment-data package.

```{r available}
available_real_timecourse_datasets()
```

Install the optional data package before running the real-data workflow:

```{r install-timecoursedata, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("timecoursedata")
```

```{r sorghum, eval=FALSE}
sorghum <- load_real_timecourse_example(
  dataset = "sorghum_leaf",
  genotype = "BT642",
  min_week = 3,
  top_n_features = 1000
)
```

For the mouse influenza microarray dataset:

```{r influenza, eval=FALSE}
flu <- load_real_timecourse_example(
  dataset = "influenza_mouse",
  group = c("C", "K", "M"),
  top_n_features = 1000
)
```

## Session Information

```{r session-info}
sessionInfo()
```
