---
title: "ModuleViz: Longitudinal Heatmap Tutorial"
author:
- name: Hua Huang
  email: helenhuang.math@gmail.com
date: "2026-05-01"
package: ModuleViz
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
abstract: |
  This tutorial introduces the main ModuleViz workflow with the bundled
  heatmap_example dataset. It covers module-number diagnostics, longitudinal
  clustering, heatmap rendering, module trajectory plots, paired omics
  visualization, pseudobulk summaries, and exportable result tables.
vignette: >
  %\VignetteIndexEntry{ModuleViz: Longitudinal Heatmap Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`r Biocpkg("ModuleViz")` clusters feature trajectories from longitudinal omics
experiments and renders ordered temporal heatmaps.

The workflow is useful for RNA-seq, ATAC-seq, CUT&RUN, or other
feature-by-sample matrices where columns represent ordered time points,
treatment conditions, or both.

This vignette uses the small bundled `heatmap_example` dataset so the tutorial
runs without network access.

# Load The Example Time-Course

```{r load}
library(ModuleViz)
data("heatmap_example")

mat <- heatmap_example$mat
meta <- heatmap_example$meta

dim(mat)
head(meta)
```

The returned object is plain R data:

- `mat`: numeric feature-by-sample matrix.
- `meta`: one row per sample, with sample ID, condition, time, and replicate.

# Choose The Number Of Modules

`choose_k()` calculates a simple within-cluster-sum-of-squares elbow diagnostic.
Use it as a first pass, then choose a biologically interpretable value.

```{r choose-k}
diag <- choose_k(
  zscore_rows(mat),
  k_range = 2:5,
  method = "kmeans"
)

diag
attr(diag, "suggested_k")
```

# Cluster Temporal Patterns

```{r cluster}
obj <- longitudinal_cluster(
  mat,
  meta,
  sample_col = "SampleID",
  time_col = "TimeHr",
  group_col = "Condition",
  k = 4,
  method = "kmeans",
  aggregate_replicates = TRUE,
  stability = TRUE,
  n_resample = 10
)

obj
```

Clusters are reordered by their temporal peak, so early-high modules appear
before late-high modules.

# Draw The Heatmap

```{r heatmap, fig.width=8, fig.height=8}
longitudinal_heatmap(
  obj,
  annotation_cols = c("Condition", "TimeHr"),
  top_n_label = 3,
  label_fontsize = 11,
  cluster_palette = "spectral",
  title = "Example longitudinal heatmap"
)
```

To save a PDF:

```{r save-heatmap, eval=FALSE}
longitudinal_heatmap(
  obj,
  annotation_cols = c("Condition", "TimeHr"),
  top_n_label = 3,
  file = "example_heatmap.pdf",
  width = 9,
  height = 12
)
```

# Plot Module Trajectories

```{r lineplot, fig.width=7, fig.height=7}
pattern_lineplot(obj, palette = "nature", base_size = 12)
```

# Draw Side-By-Side RNA And ATAC Patterns

The primary matrix drives clustering and row order. The secondary matrix is
aligned by gene and displayed next to it. This example simulates two ATAC peaks
per gene using the bundled expression data so the code is fully runnable.

```{r dual-omics, fig.width=10, fig.height=8}
set.seed(99)
genes <- rownames(mat)
atac <- mat[rep(seq_along(genes), each = 2), , drop = FALSE]
rownames(atac) <- paste0(
  "Peak",
  sprintf("%03d", seq_len(nrow(atac))),
  "_",
  rep(genes, each = 2)
)
atac <- atac + matrix(rnorm(length(atac), sd = 0.35), nrow = nrow(atac))

dual_omics_heatmap(
  primary = obj,
  secondary = atac,
  primary_name = "RNA",
  secondary_name = "ATAC",
  gene_of_primary = function(id) id,
  gene_of_secondary = function(id) sub("^.*_", "", id),
  aggregate = "mean",
  label_features = c("Gene001", "Gene016", "Gene031", "Gene046"),
  label_fontsize = 11,
  cluster_palette = "spectral"
)
```

# Single-Cell Pseudobulk Example

Single-cell data can be summarized into pseudobulk profiles by cell type,
condition, time point, and replicate. The resulting feature-by-pseudobulk matrix
uses the same ModuleViz functions.

```{r single-cell-pseudobulk, fig.width=9, fig.height=7}
cell_types <- c("T_cell", "Monocyte", "B_cell")
cells_per_sample_type <- 10
set.seed(2026)

cell_meta <- do.call(rbind, lapply(seq_len(nrow(meta)), function(i) {
  do.call(rbind, lapply(cell_types, function(ct) {
    data.frame(
      CellID = sprintf("%s_%s_Cell%03d", meta$SampleID[i], ct,
                       seq_len(cells_per_sample_type)),
      SampleID = meta$SampleID[i],
      Condition = meta$Condition[i],
      TimeHr = meta$TimeHr[i],
      Replicate = meta$Replicate[i],
      CellType = ct,
      stringsAsFactors = FALSE
    )
  }))
}))

cell_mat <- matrix(NA_real_, nrow = nrow(mat), ncol = nrow(cell_meta),
                   dimnames = list(rownames(mat), cell_meta$CellID))
cell_type_effect <- c(T_cell = 0.35, Monocyte = -0.15, B_cell = 0.05)

for (j in seq_len(nrow(cell_meta))) {
  mu <- mat[, cell_meta$SampleID[j]] + cell_type_effect[[cell_meta$CellType[j]]]
  cell_mat[, j] <- mu + rnorm(length(mu), sd = 0.55)
}

pseudobulk_group <- paste(cell_meta$CellType, cell_meta$Condition,
                          cell_meta$TimeHr, cell_meta$Replicate, sep = "|")
pseudobulk_ids <- unique(pseudobulk_group)
pseudobulk_mat <- sapply(pseudobulk_ids, function(id) {
  rowMeans(cell_mat[, pseudobulk_group == id, drop = FALSE])
})
rownames(pseudobulk_mat) <- rownames(mat)
colnames(pseudobulk_mat) <- paste0("PB", seq_along(pseudobulk_ids))

parts <- strsplit(pseudobulk_ids, "\\|")
pseudobulk_meta <- data.frame(
  SampleID = colnames(pseudobulk_mat),
  CellType = vapply(parts, `[`, character(1), 1),
  Condition = vapply(parts, `[`, character(1), 2),
  TimeHr = as.numeric(vapply(parts, `[`, character(1), 3)),
  Replicate = as.integer(vapply(parts, `[`, character(1), 4)),
  stringsAsFactors = FALSE
)
pseudobulk_meta$Group <- paste(pseudobulk_meta$CellType,
                               pseudobulk_meta$Condition, sep = "_")

sc_obj <- longitudinal_cluster(
  pseudobulk_mat,
  pseudobulk_meta,
  sample_col = "SampleID",
  time_col = "TimeHr",
  group_col = "Group",
  k = 5,
  aggregate_replicates = TRUE,
  stability = FALSE
)

pattern_lineplot(sc_obj, palette = "science", base_size = 11)
```

# Export Tables

```{r write}
out_dir <- tempdir()
write_memberships(obj, prefix = file.path(out_dir, "heatmap_example"))
```

This writes:

- `heatmap_example_memberships.txt`
- `heatmap_example_centroids.txt`
- `heatmap_example_stability.txt`

# Real Public Data

The package also includes a helper for public datasets in Bioconductor
`timecoursedata`. These examples are optional because they require installing
the experiment-data package.

```{r available}
available_real_timecourse_datasets()
```

Install the optional data package before running the real-data workflow:

```{r install-timecoursedata, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("timecoursedata")
```

```{r sorghum, eval=FALSE}
sorghum <- load_real_timecourse_example(
  dataset = "sorghum_leaf",
  genotype = "BT642",
  min_week = 3,
  top_n_features = 1000
)
```

For the mouse influenza microarray dataset:

```{r influenza, eval=FALSE}
flu <- load_real_timecourse_example(
  dataset = "influenza_mouse",
  group = c("C", "K", "M"),
  top_n_features = 1000
)
```

# Session Information

```{r session-info}
sessionInfo()
```
