---
title: "ModuleViz: temporal modules from longitudinal omics"
author:
- name: Helen Huang
  email: helenhuang.math@gmail.com
package: ModuleViz
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
abstract: |
  ModuleViz turns a feature-by-sample matrix with an ordered sample axis into
  interpretable temporal modules: groups of genes or peaks that rise and fall
  together. It clusters trajectories, orders the modules from earliest-peaking
  to latest-peaking, measures how reproducible each module is, and renders the
  result as publication-ready heatmaps, trajectory plots, and per-module
  correlation maps. Paired omics layers such as RNA and CUT&Run can be aligned
  by gene and shown side by side. This vignette is the complete user guide.
vignette: >
  %\VignetteIndexEntry{ModuleViz: temporal modules from longitudinal omics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dpi = 110,
  fig.align = "center"
)
set.seed(1)
```

# Introduction

## What problem does this solve?

A longitudinal omics experiment produces a matrix of features (genes, peaks,
proteins) measured across samples that have a meaningful **order**: time points,
doses, disease grades, or differentiation stages. The interesting signal is
rarely in any single feature. It is in the *shapes*: which features rise early
and fall, which come up late and stay up, which respond only in the treated arm.

`ModuleViz` finds those shapes and makes them legible. The workflow is:

1. **z-score** each feature so shape matters, not absolute abundance;
2. **cluster** features into temporal modules;
3. **order** the modules from earliest-peaking to latest-peaking, so the heatmap
   reads like a timeline from top to bottom;
4. **assess** how reproducible each module is under resampling;
5. **visualise** modules as heatmaps, mean-trajectory line plots, and
   gene-gene correlation maps;
6. optionally **align a second omics layer** by gene.

## When to use it

`ModuleViz` fits any experiment where the columns have an order and the rows are
features:

* bulk RNA-seq / ATAC-seq / CUT&Run time courses;
* drug or stimulation time courses, with or without a control arm;
* differentiation or disease-stage series (the axis need not be clock time);
* pseudobulk profiles derived from single-cell data.

It deliberately does *not* do feature selection or differential testing. Choose
which features are interesting upstream — with `r Biocpkg("DESeq2")`,
`r Biocpkg("edgeR")`, `r Biocpkg("limma")`, or a variance filter — and pass in
the matrix you want to see.

## The two vignettes

| Vignette | Data | Purpose |
|---|---|---|
| **This one** | Bundled synthetic dataset | Complete API reference, runs offline |
| `vignette("cd8-exhaustion-published-data")` | GEO GSE285248 + GSE285245 | Real published RNA-seq + CUT&Run walkthrough |

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

# then install ModuleViz itself, e.g.
# remotes::install_github("helenhuangmath/ModuleViz")
```

```{r load-pkg, message = FALSE}
library(ModuleViz)
```

# Quick start

Five lines from data to figure. Everything after this section explains what
these calls are doing and how to control them.

```{r quickstart, fig.width = 8, fig.height = 7, fig.cap = "Quick-start heatmap. Rows are split into four temporal modules ordered early-peaking to late-peaking; columns are split by condition and ordered by time."}
data("heatmap_example")

obj <- longitudinal_cluster(
    heatmap_example$mat, heatmap_example$meta,
    sample_col = "SampleID", time_col = "TimeHr", group_col = "Condition",
    k = 4, aggregate_replicates = TRUE
)

longitudinal_heatmap(obj, top_n_label = 3, title = "Quick start")
```

# Input data

## The two objects you supply

`ModuleViz` needs exactly two things.

**A numeric matrix**, features in rows, samples in columns, with feature IDs as
`rownames()`:

```{r input-mat}
mat <- heatmap_example$mat
dim(mat)
mat[1:4, 1:4]
```

**A metadata table**, one row per sample:

```{r input-meta}
meta <- heatmap_example$meta
head(meta, 4)
```

Three columns in the metadata have special meaning, named via arguments:

```{r input-args, echo = FALSE}
knitr::kable(data.frame(
  Argument = c("`sample_col`", "`time_col`", "`group_col`"),
  Required = c("yes", "yes", "no"),
  Meaning = c(
    "Sample IDs; must match the matrix column names.",
    "**Numeric** ordering variable. Sorted ascending to order the columns.",
    "Optional second factor. Splits heatmap columns and colours the line plot."
  )
))
```

The critical constraint is that `time_col` must be **numeric**. It is what
orders the columns and what defines a module's peak. If your axis is
categorical but ordered — `"naive" < "memory" < "exhausted"`, or
`"low" < "mid" < "high"` — encode it as an integer rank:

```{r ordinal, eval = FALSE}
meta$StageOrder <- c(naive = 1, memory = 2, exhausted = 3)[meta$Stage]
```

Samples present in the metadata but missing from the matrix (or vice versa) are
silently dropped, so the two only need to overlap.

## Feature-level preprocessing

Cluster on a **log or variance-stabilised scale** for count data. Z-scoring raw
counts lets a handful of very high-count features dominate the distance metric:

```{r preprocess, eval = FALSE}
mat <- log2(counts + 1)                       # or DESeq2::vst(dds)
mat <- mat[rowMeans(counts) >= 10, ]          # drop low-expressed features
```

# The core workflow

## Z-scoring

`zscore_rows()` centres and scales each feature across its samples, so that a
gene expressed at 10 counts and one at 10,000 counts contribute equally if they
have the same shape. Zero-variance rows become `0` rather than `NaN`, so they
survive clustering instead of breaking it.

```{r zscore}
z <- zscore_rows(mat)
round(rowMeans(z)[1:3], 12)     # centred
round(apply(z, 1, sd)[1:3], 12) # scaled
```

`longitudinal_cluster()` calls this internally. Pass `is_zscore = TRUE` if your
matrix is already z-scored.

## Choosing the number of modules

`choose_k()` scans a range of `k` and reports two diagnostics: the total
within-cluster sum of squares (an elbow curve) and the mean silhouette width.

```{r choose-k, fig.width = 5.5, fig.height = 4}
diag <- choose_k(z, k_range = 2:8, method = "kmeans")
diag
attr(diag, "suggested_k")
```

When you settle on a `k` that differs from the elbow, pass it back as
`chosen_k`: the diagnostic figure then draws your `k` as a solid line beside the
dashed elbow, so it stays consistent with the module count in the rest of your
figures.

```{r choose-k-marked, eval = FALSE}
choose_k(z, k_range = 2:8, chosen_k = 4, file = "elbow.pdf")   # mark it
choose_k(z, k_range = 2:8, annotate_k = FALSE, file = "elbow.pdf")  # bare curve
```

The suggested `k` is the point furthest from the straight line joining the first
and last elbow points. **Treat it as a starting point, not an answer.** The
right `k` is the one that separates programmes you can interpret. Over-splitting
is usually more informative than under-splitting: two near-identical modules are
easy to merge mentally, whereas one module hiding two anti-correlated programmes
is actively misleading (see Section \@ref(module-coherence)). The real-data
vignette works through a concrete case where the elbow k merges two biologically
distinct programmes and the reported k is deliberately one higher.

## Clustering

`longitudinal_cluster()` is the workhorse.

```{r cluster}
obj <- longitudinal_cluster(
    mat, meta,
    sample_col = "SampleID",
    time_col   = "TimeHr",
    group_col  = "Condition",
    k          = 4,
    method     = "kmeans",
    aggregate_replicates = TRUE,
    stability  = TRUE,
    n_resample = 20,
    seed       = 1
)
obj
```

### Clustering methods

```{r methods-table, echo = FALSE}
knitr::kable(data.frame(
  `method` = c("`\"kmeans\"`", "`\"hierarchical\"`", "`\"pam\"`"),
  Notes = c(
    "Default. Fast, Hartigan-Wong, `nstart = 50` random starts.",
    "Ward.D2 linkage by default via `hclust_method`. Deterministic.",
    "Partitioning around medoids; more robust to outliers than k-means."
  ),
  check.names = FALSE
))
```

`dist_method` accepts any `stats::dist()` metric plus `"correlation"`
(`1 - Pearson r`), which is often the better choice for trajectory shape because
it ignores amplitude entirely:

```{r alt-method}
obj_corr <- longitudinal_cluster(
    mat, meta, sample_col = "SampleID", time_col = "TimeHr",
    k = 4, method = "hierarchical", dist_method = "correlation",
    stability = FALSE
)
table(obj_corr$cluster)
```

### Replicate aggregation

`aggregate_replicates = TRUE` averages replicate columns within each
`group_col` × `time_col` cell before clustering. This gives cleaner modules and
a much narrower heatmap, at the cost of no longer seeing replicate variability.
Keep it `FALSE` when you want to inspect reproducibility across replicates.

## What you get back

The result is a `longi` object — a plain list, so everything is inspectable:

```{r longi-structure}
names(obj)
```

```{r longi-table, echo = FALSE}
knitr::kable(data.frame(
  Element = c("`z`", "`cluster`", "`membership`", "`centroids`", "`stability`",
              "`meta`", "`params`"),
  Contents = c(
    "z-scored matrix, rows reordered by module then signal strength.",
    "Ordered factor giving each row's module.",
    "`data.table` of feature ID, module, and max |z|.",
    "Mean z per module per (group x) time point.",
    "Mean Jaccard per module, or `NULL`.",
    "Sample metadata, reordered to match the matrix columns.",
    "The settings used, for reproducibility."
  )
))
```

```{r membership}
head(obj$membership, 4)
```

## How modules get ordered {#ordering}

This is the feature that makes the heatmap readable, so it is worth stating
explicitly. After clustering, each module's centroid is computed across the
ordered time points. Modules are then ranked by **where that centroid peaks**
(its arg-max), with ties broken by overall slope so that, among modules peaking
at the same time, rising ones come before falling ones.

The consequence: `Cluster1` always peaks earliest and `ClusterK` latest.

```{r centroids}
cent <- as.matrix(obj$centroids[, -1])
rownames(cent) <- obj$centroids$Cluster
round(cent, 2)
```

Within a module, rows are sorted by max |z|, so the strongest features sit at
the top of each block.

## Cluster stability

A module you cannot reproduce is not a finding. `cluster_stability()` — run
automatically when `stability = TRUE` — repeatedly subsamples the features,
re-clusters, and records how well each original module is recovered, as a mean
maximum Jaccard similarity.

```{r stability}
obj$stability
```

```{r stability-guide, echo = FALSE}
knitr::kable(data.frame(
  `Mean Jaccard` = c("> 0.85", "0.6 - 0.85", "< 0.6"),
  Interpretation = c("Highly reproducible; report with confidence.",
                     "A real pattern, but boundaries are fuzzy.",
                     "Unstable. Do not interpret as a discrete module."),
  check.names = FALSE
))
```

Stability is shown twice in the heatmap: as a grey `Stability` bar down the left
edge, and as `J=` in each row title.

# Visualisation

## The heatmap

```{r heatmap-main, fig.width = 8.5, fig.height = 7.5, fig.cap = "Modules ordered early to late, columns split by condition, top three features labelled per module."}
longitudinal_heatmap(
    obj,
    annotation_cols = c("Condition", "TimeHr"),
    top_n_label     = 3,
    title           = "Stimulation time course"
)
```

Reading the figure: rows are split by module and never re-ordered by the
heatmap itself (`cluster_rows = FALSE`), because the ordering from Section
\@ref(ordering) is the point. Columns are split by `group_col` and ordered by
time. Metadata columns become annotation bars along the top.

### Labelling features

Three ways to choose which features get labelled, combinable:

```{r labels, eval = FALSE}
# 1. an explicit list
longitudinal_heatmap(obj, label_features = c("Gene001", "Gene016"))

# 2. top N per module by signal strength
longitudinal_heatmap(obj, top_n_label = 5)

# 3. top N per module by YOUR statistic (e.g. -log10 padj from DESeq2)
rank_tab <- data.frame(ID = rownames(obj$z), value = runif(nrow(obj$z)))
longitudinal_heatmap(obj, top_n_label = 5, rank_table = rank_tab,
                     id_field = "ID", value_field = "value")
```

Option 3 is usually what you want for a paper: it labels the features that are
most *significant*, not merely the most variable.

### Saving

Pass `file` to write directly. The extension picks the device — `.pdf` for
vector output, `.png` for a raster at 300 dpi.

```{r save, eval = FALSE}
longitudinal_heatmap(obj, file = "modules.pdf", width = 9, height = 12)
```

## Palettes {#palettes}

Three independent palette slots, each with its own job:

```{r palette-table, echo = FALSE}
knitr::kable(data.frame(
  Argument = c("`cluster_palette`", "`annotation_palette`", "`zscore_palette`"),
  Colours = c("module identity bar", "categorical metadata bars", "the matrix body"),
  Default = c("`\"paired\"`", "`\"set2\"`", "`\"rdbu\"`"),
  Options = c("any qualitative or ramp palette below",
              "any qualitative palette below",
              "`\"rdbu\"`, `\"viridis\"`")
))
```

The registry distinguishes two kinds of palette, and this distinction matters:

* **Qualitative** — `"paired"`, `"set2"`, `"set3"`, `"okabe_ito"`, `"nature"`,
  `"science"`. Hand-picked, maximally distinguishable colours. For `n` at or
  below the palette size, the first `n` are used **verbatim**. Interpolating
  them would invent blended colours that are not in the palette and can be
  nearly identical to each other.
* **Continuous ramps** — `"spectral"`, `"viridis"`. Genuine colour scales,
  interpolated across the full range for any `n`.

```{r palette-demo, fig.width = 7, fig.height = 3.2, fig.cap = "Qualitative palettes keep their colours verbatim; ramps interpolate across the full range."}
pals <- c("paired", "set2", "set3", "okabe_ito", "spectral", "viridis")
op <- par(mar = c(1, 6, 1, 1))
plot(NA, xlim = c(0, 6), ylim = c(0, length(pals)), axes = FALSE,
     xlab = "", ylab = "")
for (i in seq_along(pals)) {
    cols <- ModuleViz:::.moduleviz_discrete_palette(6, pals[i])
    rect(0:5, length(pals) - i, 1:6, length(pals) - i + 0.8,
         col = cols, border = "white", lwd = 2)
}
axis(2, at = seq(length(pals) - 0.6, by = -1, length.out = length(pals)),
     labels = pals, las = 1, tick = FALSE, cex.axis = 0.85)
par(op)
```

Use `viridis` when the figure must survive greyscale printing or colour-vision
deficiency; use `rdbu` when the zero point is meaningful and you want direction
to read at a glance.

```{r heatmap-viridis, fig.width = 8, fig.height = 7, fig.cap = "The same modules with a viridis body, Spectral module bar, and Set3 metadata bars."}
longitudinal_heatmap(
    obj,
    zscore_palette     = "viridis",
    cluster_palette    = "spectral",
    annotation_palette = "set3",
    top_n_label        = 2,
    title              = "Viridis variant"
)
```

`zscore_limits` controls where the scale saturates; widen it if `±2` clips too
much of your data.

## Module trajectories

The heatmap shows every feature; the line plot shows each module's average
shape, which is what you actually describe in the text.

```{r lineplot, fig.width = 6, fig.height = 7, fig.cap = "Mean z-score trajectory per module. With a group column, each group is a separate coloured line."}
pattern_lineplot(obj, palette = "set2", base_size = 12)
```

Returns a `ggplot`, so it composes with the rest of the ggplot ecosystem:

```{r lineplot-custom, eval = FALSE}
library(ggplot2)
pattern_lineplot(obj) + labs(subtitle = "Stimulated vs control")
```

## Module coherence {#module-coherence}

Clustering guarantees that a module's members are *closer to each other than to
other modules*. It does not guarantee they actually agree. A module can look
clean in the main heatmap while being the average of two anti-correlated
sub-programmes.

`module_correlation_summary()` gives the one-number version:

```{r cor-summary}
module_correlation_summary(obj)
```

```{r cor-heatmap, fig.width = 6.5, fig.height = 6, fig.cap = "Gene-gene correlation within one module. A coherent module is a solid high-correlation block; visible sub-blocks mean the module contains more than one programme."}
cors <- module_correlation_heatmap(obj, clusters = "Cluster1", max_genes = 40)
cors$Cluster1$mean_cor
```

Interpretation: high mean `r` (roughly > 0.7) means a tight, coherent module.
A low value with visible blocks in the heatmap means you should raise `k` and
re-cluster. A low value with no structure means the module is mostly noise.

Large modules are subsampled to `max_genes` for legibility —
`select = "top"` keeps the strongest-signal features, `select = "random"` gives
an unbiased view. To write every module at once, pass a `.pdf` path and get one
page per module:

```{r cor-save, eval = FALSE}
module_correlation_heatmap(obj, file = "module_correlations.pdf",
                           max_genes = 60)
```

# Paired omics

`dual_omics_heatmap()` puts two layers side by side, linked by gene: RNA next to
ATAC, or RNA next to CUT&Run. The **primary** layer drives the clustering and
the row order; the secondary is displayed in that same order, so each row is one
gene across both assays.

Because one gene often maps to several peaks, the secondary layer is collapsed
to one row per gene — `aggregate = "mean"` averages a gene's peaks,
`aggregate = "top"` takes the strongest.

Genes are recovered from secondary row IDs by `gene_of_secondary`, whose default
`sub("^.*_", "", id)` takes everything after the last underscore. So row names
of the form `peak1243_Pdcd1` work out of the box.

```{r dual-omics, fig.width = 10, fig.height = 7, fig.cap = "Paired layers linked by gene. The primary layer sets the module structure and row order; the secondary is shown in the same order."}
# simulate a second layer: two peaks per gene, with noise
genes <- rownames(mat)
atac  <- mat[rep(seq_along(genes), each = 2), ]
rownames(atac) <- paste0("Peak", sprintf("%03d", seq_len(nrow(atac))),
                         "_", rep(genes, each = 2))
atac <- atac + matrix(rnorm(length(atac), sd = 0.35), nrow = nrow(atac))

res <- dual_omics_heatmap(
    primary        = obj,
    secondary      = atac,
    primary_name   = "RNA",
    secondary_name = "ATAC",
    aggregate      = "mean",
    label_features = c("Gene001", "Gene016", "Gene031")
)
```

The return value carries the aligned matrices for downstream quantification:

```{r dual-return}
names(res)
dim(res$z_secondary)
```

That makes it straightforward to ask whether the second layer tracks the first —
for example, comparing mean secondary signal at late versus early time points
for the late-peaking module.

# Exporting results

```{r export}
demo_prefix <- file.path(tempdir(), "demo")
write_memberships(obj, prefix = demo_prefix)
list.files(tempdir(), pattern = "^demo_")
```

`_memberships.txt` is the file you feed into GO or GSEA enrichment, one module
at a time. `_centroids.txt` holds the mean trajectories, and `_stability.txt`
the Jaccard scores.

# The one-call wrapper

`run_longitudinal()` runs cluster → heatmap → line plot → export in a single
call, for when you have settled on parameters and just want the outputs:

```{r wrapper, eval = FALSE}
run_longitudinal(
    mat, meta, out_prefix = "results/experiment1",
    sample_col = "SampleID", time_col = "TimeHr", group_col = "Condition",
    k = 4, top_n_label = 5,
    zscore_palette = "rdbu", cluster_palette = "paired"
)
```

# Working with published data

`vignette("cd8-exhaustion-published-data", package = "ModuleViz")` applies this
workflow end to end to two public GEO series — RNA-seq (GSE285248) and paired
H3K27ac / H3K27me3 CUT&Run (GSE285245) from a CD8 T cell memory-versus-exhaustion
study. It covers the parts this vignette does not: downloading processed
matrices from GEO, ANOVA-based DEG selection, mapping CUT&Run peaks to genes,
and checking module assignments against canonical marker genes.

The package also ships `examples/validate_real_data.R`, which runs that entire
analysis with 114 correctness assertions.

For a small set of bundled public time courses:

```{r real-data}
available_real_timecourse_datasets()
```

# Frequently asked questions

**`Error: mat has no rownames`** — feature IDs must be `rownames(mat)`, or
passed via `id_col` if they live in a column of a `data.frame`.

**`Error: No overlap between meta[[sample_col]] and the matrix columns`** —
`meta[[sample_col]]` must contain the literal matrix column names. Check for
whitespace or a factor that has been coerced to integers.

**My modules are dominated by a few features.** You are clustering on a raw
count scale. Use `log2(x + 1)` or a variance-stabilising transform first.

**Everything lands in one module.** Usually too few features, or features that
were not selected for variability. Filter to differential or high-variance
features first.

**A module has low stability.** Lower `k`, switch to
`dist_method = "correlation"`, or accept that the module is a continuum rather
than a discrete group and describe it that way.

**Can the axis be non-numeric?** No — `time_col` must be numeric. Encode ordered
categories as integer ranks; the spacing does not need to be even.

**Can I use my own colours?** Yes. `annotation_colors` takes a named list of
colour vectors, and `col_fun` takes any `circlize::colorRamp2()` function,
overriding `zscore_palette`.

# Session info

```{r session-info}
sessionInfo()
```
